Prediction of Consulting Project Profitability

### 2016-12-13 finalwriteup/Consulting Profitability Paper/test.tex
reset

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Predictive Models to Support Quoting of Fixed Fee Consulting Projects},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  \title{Predictive Models to Support Quoting of Fixed Fee Consulting Projects}
    \author[Queensland University of Technology]{Amy Cook, Paul Wu, Kerrie Mengersen\corref{c1}}
   \ead{a21.cook@qut.edu.au} 
   \cortext[c1]{Corresponding Author}
      \address[Queensland University of Technology]{School of Mathematical Sciences, George Street, Brisbane, QLD, 4000}

@@

Clearly state the research question and objectives of the work. Briefly
provide any necessary background to frame the research question.
Concisely summarize the major findings/results.

Depending on the industry, some consulting projects can be financialy
risky endeavors for the consultant. Engineering consultanting companies
in the construction industry complete a wide range of complex projects
in a competitive enviroment. Because of the way contracts are
structured, many of these projects result in losses. The purpose of this
study was to test whether these loss making projects could be predicted
by statistical and machine learning algorithms using the information
available at project conception. The second question is whether the
predictive algorithm could improve a company's profitability. Project
data from a case study engineering company was used to answer these
questions.

@@

@@

Financial risk most commonly falls upon a consulting company when they
must offer a fixed fee for a project. In competitive climates, clients
request fixed price quotes for project completion from several
consultants before a project is awarded. Hence, initial quoted fees must
be competitive but also accurately cover consultant salaries and
business costs.

How do consultants make this accurate yet competitive fee estimate? This
varies from industry to industry, but typically a consulting manager has
experience in the type of project they are quoting and, after reviewing
the project details, can use a combination of intuition, comparison to
past projects, and rules of thumb. Unfortunately, some projects priced
using this method can blow out of budget, causing significant losses to
the consultant.

A review of 6 IT project surveys between 1984 and 1994 by Moløkken \&
Jørgensen (2003) revealed that 60-80\% of IT projects encountered effort
and/or schedule overruns, where the average overrun was by 30 to 40\%. A
study on large-scale infrastructure projects over the past seventy years
revealed that cost forecasts consistently underestimated the cost of
rail projects by an average of 44.7\%, the cost of bridge and tunnel
projects by 33.8\% and the cost of road projects by 20.4\% (Flyvbjerg
2007).

Kahneman won a nobel prize in 2002 for his research into decision making
on this topic. He argued that a person's natural optimistic view of
their own skills leads to consistent underestimation of the time and
risks involved in a project (Lovallo and Kahneman 2003). To counter this
phenomenon, research by Flyvbjerg {[}-Flyvbjerg2011{]} advised that
managers should be confronted by the financial performance of similar
projects to the one being costed so that they take on a more realistic
view of their abilities. This was based off a body of findings where
people made better predictions about themselves after being exposed to
facts about other people. For example, subjects were asked to predict
their skill levels before and after being exposed to a summary of other
people's skills, and their predictions after were significantly more
accurate (Lovallo and Kahneman 2003).

This work aims to provide a manager with this more realistic view by
predicting whether the new project will be profitable or not, based on a
suite of previous project data from the same company. The manager could
then take multiple actions including not tendering for a project
predicted to be loss-making, providing a high fixed price, or
restructuring the contract so that it is not 100\% fixed price.

@@

This thesis focuses on a single case study consulting company in the
construction industry. twelve years of passively collected data that
described each project in terms of their clients, invoice history,
employee hours and technical details. labour is the chief cost and the
long traditional history of this industry mandates fixed price projects
as the norm. 20\% of projects are loss making. Fees calculated via
detailed analysis A Customer Relationship Management software package
(CRM) is currently employed to collect and store project data. A CRM is
a popular type of software used by businesses to record client and
project details as well as employee time sheet records. In the case
study business, the CRM is available to all employees over the company
intranet, and each employee completes daily time sheets allotting their
hours to certain projects. Additionally, technical information is
recorded against each project

\begin{itemize}
\tightlist
\item
  Employee time sheet hours with dates
\item
  Other project costs (taxis, printing)
\item
  Client information/characteristics
\item
  Client identification code
\item
  Invoiced amounts for each project and dates
\item
  Employee costs
\item
  Employee charge out rates
\item
  Project description
\end{itemize}

2364 past projects are available. missing data due to how the data
collection changed over time and some fields were not mandatory.

To assess whether the model would improve the case study company's
profitability, a simple managerial decision framework was applied: if
the algorithm predicted the project to be loss making, the project was
considered `rejected' and all profits or \emph{losses} of rejected
projects were discarded. The remaining projects' profits and losses were
summed to give a revised overall profit over the last 12 years of
projects. This was compared to the summed profits and losses from all
projects completed over the last 12 years. The best algorithm rejected
enough loss-making projects to improve overall profits by 9\%.

\begin{itemize}
\tightlist
\item
  aim \textbf{General Aim} Use statistical techniques to model the
  profitability of projects for consulting businesses using their
  internal CRM data. Research will focus on a case study Engineering
  consulting company that offers their expert advice (in the currency of
  time) to business clients. The project outcomes are intended to assist
  the business in predicting project profitability before engagement.
  Several statistical and machine learning techniques were tested,
  compared and refined.
\end{itemize}

\textbf{Hypothesis 1} A statistical or machine learning model based on
historical project data can predict the profitability of a new project
with greater accuracy than a baseline predictor. A baseline predictor
for this project is one that predicts the average of a numeric response
variable for all cases or, if the response variable is categorical,
randomly assigned categories for each case, where the proportion of
assigned categories matches the true categorical proportions.

\textbf{Hypothesis 2} The predictive model built from Hypothesis 1 can
be shown to have a positive impact on the overall profit earned by the
case study business. The overall profit is represented by the following
equation:

@@

The bulk of research to date has been performed with project data in
either the construction industry or software development. Research in
the construction industry has primarily focused on predicting the final
building cost of construction. Over the past few decades, significant
research has been dedicated to creating predictive models that present
the `outside view' of a project. This was done by statistically
comparing a new project to a collection of similar projects and their
characteristics. It has been shown that these models improve cost
estimation accuracy, however industry has not adopted them.

\subparagraph{2.1.1 Construction Industry}\label{construction-industry}

Elfaki, Alatawi, and Abushandi (2014) completed a review of cost
estimation research from 2004 to 2014. They found that artificial Neural
Networks and SVM's were the most common machine learning techniques.
Neural Networks and SVM's received significant attention in the 1990's
for their ability to accurately predict construction costs with limited
detailed information (Shin 2015; Kim, An, and Kang 2004). Shin (2015)
pioneered the application of Boosted Trees to cost estimation in
construction projects. Often, Linear Regression was the only model
assessed, without comparison to other methods such as Neural Networks,
which first started appearing in literature in the 1990's (Kim, An, and
Kang 2004) Some studies showed that Neural Networks outperform
Regression, however other studies established they are approximately
equal (Kim, An, and Kang 2004; Attalla and Hegazy 2003).

Variety of variables used: (Chan and Park 2005)(Akintoye and Fitzgerald
2000; Trost and Oberlender 2003; Pinto and Slevin 1988) the studies all
generally reported their mean absolute error or a similar metric from
the tested models, and results were generally positive. Furthermore, the
existing studies collected data from many businesses for a single study

\subparagraph{2.1.2 Software Industry}\label{software-industry}

In the software industry, the main component of cost is \emph{effort} as
opposed to the cost of building materials in the construction industry.

Similar to the construction industry, expert judgment or detailed
analysis, is the most widely practiced method for effort estimation
(Shepperd, Schofield, and Kitchenham 1996; Moløkken and Jørgensen 2003)

Multiple studies have shown that Neural Networks definitively outperform
regression models in effort estimation, although Regression is the most
popular method (Finnie, Wittig, and Desharnais 1997; Pai, McFall, and
Subramanian 2013; Matson and Mellichamp 1993).

Interestingly, several studies have found that even if 15 or so
variables are included, often only one variable contributes
significantly to the model's accuracy: size (or funciton point)
(Shepperd, Schofield, and Kitchenham 1996; Finnie, Wittig, and
Desharnais 1997; Pai, McFall, and Subramanian 2013)

A study by Mendes and Kitchenham (2004) demonstrated using 67 web
projects that cross-company models were significantly less accurate than
a within company model (Shepperd, Schofield, and Kitchenham 1996) Again,
Neural Networks were criticised for their inability to explain their
results (Finnie, Wittig, and Desharnais 1997). The software industry is
unique in that agile methods of delivery are changing the contractual
approach of consulting (Badenfelt 2011). Software development is a
relatively new field in comparison to construction, where the tradition
of fixed fee structures is difficult to reinvent.

@@

The review of methods applied to the construction and IT industry
highlighted the use of Linear Regression, Neural Networks, SVM's and in
one case Boosted Trees, however research on other business problems
utilised a wider range of methods. These included Naive Bayes, Random
Forests, and machine learned Bayesian Networks.

Although SVM's performed the best, as discussed, they are often complex
and slow, requiring a great deal of memory (Kumar and Ravi 2007). Kumar
and Ravi (2007) performed a detailed review of statistical and machine
learning techniques that were applied over 37 years in the context of
bankruptcy prediction in banks.

For example in financial credit scoring, a study by Brown and Mues
(2012) found that Random Forests and Boosted Trees consistently
outperformed Neural Networks in classification. Kumar and Ravi's (2007)
review also assessed ensemble techniques, which refers to combinations
of two completely different algorithms, and found they can often
outperform individual methods.

An exception is Saradhi \& Palshikar's (2011) study on employee churn,
where `churn' refers to the number of individuals moving out of a group
within a certain time. Naive Bayes, Random forests (an ensemble decision
tree method) and SVM's were built. determining the value of each
employee in terms of the importance of the projects they were on and
their monthly chargeability. This allowed them to rank employees
identified as `high risk of churn' by value and provided a clear ranking
for manager's to act upon. This extension of the study provided a
comprehensive framework for how business managers could adopt their
findings to improve business operations. It was a valuable addition that
is absent from most cost and effort estimation research.

\paragraph{2.3 Gaps}\label{gaps}

In comparison to previous studies, this project advances the body of
work on cost estimation in three ways: the data set is larger than other
studies and sourced from a single company's CRM database, ensemble tree
methods and model blending was tested, and the prediction results were
integrated into a decision framework for the business. In past research,
cost estimation data sets generally consisted of \textless{}300 of
projects from a range of a companies and even countries. This thesis
case study only had access to the company's internal data but this
contained over 2,000 past jobs Furthermore, the problem of cost
estimation is applied to a consulting company in the construction
industry, which differs from other studies in the construction industry
that estimate actual construction costs. Consulting in the construction
industry can be likened more to software development effort estimation
although the nature of the work differs considerably.

ensemble tree methods and blended methods are applied which have been
minimally tested on this problem. To date, Linear Regression, Neural
Networks, Case Based Reasoning (CBR), and Support Vector Machines
(SVM's) have been used to predict cost/effort even though ensemble tree
methods can perform as well as Neural Networks and generally outperform
Linear Regression (Caruana and Niculescu-Mizil 2006). Another benefit of
ensemble trees is the output types available, such as partial dependency
plots and variable importance plots that provide insight into the
model's calculations. This contrasts Neural Networks and SVM's, which
are `black-box' predictors.

Finally, this study propels the predictive model one step further than
other studies by analysing bottom-line profit improvements that could
result from the research. final profit increases are evaluated. This
kind of thought experiment is designed to present a clear case for
industry when evaluating if and how to adopt cost estimation models.

\section{3. Prediction Methods}\label{prediction-methods}

@@

Neural Netowrks and SVM were not purusued because black box A previously
mentioned disadvantage is that the NN algorithm is a black box because
the internal structure is too complex for interpretation. They also
require a lot of training data relative to other methods.

Also, SVM's can be very slow to train and therefore not suitable for
industry purposes (Auria and Moro 2008). Because it is so important to
engage decision makers with a model that can explain its results, these
two methods are not appropriate for the effort estimation problem.

@@

binary classification problems, where the objective is to predict the
probability of an event occurring. Probability is a continuous response
variable, however predictions from a linear regression would not be
bound by 0 and 1.

\subparagraph{3.1.1 Logistic Regression}\label{logistic-regression}

Because of the linear relationship, an equation can be fit to the log
odds of the binary response variable against each of the explanatory
variables using linear regression. Then, the linear equation describing
the log odds can be transformed back to probability by taking the
inverse log i.e.~the exponential and rearranging (Macdonald 1975):

The result in a sigmoidal function bound by 0 and 1. The sigmoidal
function originates from a linear fit of the log odds in the data.
Assume linear relationship between covariates and response variable

A coefficient represents the change in the log odds of the response
variable for each unit increase in an explanatory variable. Therefore,
taking the exponential of the coefficient is the change in odds of the
response variable. If the change in odds was 2, then if the explanatory
variable increased by 1, the response variable event would be twice as
likely to occur.

Logistic Regression is a good benchmark to compare other binary
predictive models due to its simplicity and speed (Moore and McCabe
1989). single linear relationship means Not prone to overfitting

\subparagraph{3.1.2 Random Forests}\label{random-forests}

single decision trees also suffer from low predictive accuracy and
instability, so to combat these problems, ensemble tree methods were
pioneered in the 1990's (Breiman 1996).

The three methods of combining hundreds or thousands of trees turn
decision trees into high performing, stable predictors.

training multiple trees from the same data by sampling bootstrapped
training sets with replacement. However, when creating each tree, a
random subset of attributes (variables) is considered at each split. The
reduced subset of attributes is resampled for each split in the tree.
This allows dominant variables to be suppressed for a fraction of the
splits, allowing the algorithm to explore signals in weaker variables
(Breiman 2001).

Additionally, ensemble tree methods are faster than SVM's and Neural
Networks but can perform just as well and even provide insights such as
variable importance and variable relationships (Sealfon and Gymrek
2012).

\subparagraph{3.1.3 Gradient Boosted
Trees}\label{gradient-boosted-trees}

Lastly, the boosted decision tree approach applies gradient descent
theory to a series of decision trees. The trees are limited to a certain
depth to maintain simplicity, and each tree models the residuals (or
errors) of the preceding tree. By modeling the errors, misclassified
cases are weighted higher than correctly classified cases and influence
the structure of the latest tree more (Hastie, Tibshirani, and Friedman
2009). This increased weighting on errors is why the algorithm is called
boosting. The limited depth of each tree prevents overfitting at each
stage and the combined result of up to thousands of trees is very
powerful (Elith, Leathwick, and Hastie 2008).

Caruana and Niculescu-Mizil (2006) tested Boosted Trees, Random Forests,
Neural Networks, SVM's, Logistic Regression and Naive Bayes on 11 binary
classification problems and found that Boosted Trees performed best,
followed by Random Forests. This demonstrates how ensemble tree methods
are capable of competing with high-level machine learning algorithms.

Trees are not built on a probabilistic framework, and therefore their
results cannot be provided in this framework (Louppe and Prettenhofer
2014).

\subparagraph{3.1.4 Naive Bayes}\label{naive-bayes}

The Naive Bayes method works by making conditional independence
assumptions about the explanatory variables in order to simplify
probability calculations for the response variable (the response
variable must be categorical).

These equations are simple to compute given the data, and the class with
the highest probability can then be chosen (F. Provost and Fawcett
2013).

The advantages of this method are that the conditional independence
assumption enables very fast calculations and predictions. The method
can perform well in real world tasks because the assumption of
independence does not significantly damage predictions (F. Provost and
Fawcett 2013). Assumptions of independence genearlly not true
(correlated variables). This is fine for ranking but the output
probabilities are not accurate statistical probabilities (Caruana and
Niculescu-Mizil 2006).

Naive Bayes method generally provides a good benchmark to compare
against more complex models that should outperform it (Caruana and
Niculescu-Mizil 2006).

\subparagraph{3.1.5 Bayesian Networks}\label{bayesian-networks}

A Bayesian Network is a graphical probabilistic model that illustrates
the conditional dependencies between variables in a data set. The model
is visually represented by a directed acyclic graph (DAG) and is capable
of linking the conditional dependency between any variable to another
variable (Heckerman 1998). The conditional relationships are Bayesian,
where the probabilities in one node are conditional upon values in nodes
directed towards it as well as preceding nodes

Bayesian Networks reduce computations required to find the probability
of a unique combination given other explanatory variable values while
still taking into account many conditional dependencies that Naive Bayes
(Barber 2012)

Naive Bayes is the simplest form of a Bayesian Network (Zhang 2004)

Network relationships can be learned from the data, however this is not
widely included as part of the suite of machine learning methods. The
conditional variable dependencies are calculated from the data, which in
turn can define the graph structure. Some conditional dependencies can
be set before the structure is learned (Barber 2012).

Drawbacks of Bayesian Networks include the difficulty for them to
process continuous variables.

Bayesian networks have found success in combining deterministic models
with observed data as well as expert knowledge.(Kragt 2009)

@@

It includes first how the data was obtained, the lengthy cleaning
process, Projects varied from total invoiced amounts of \$500 to over
\$1,000,000 a project could have thousands of rows of relevant data that
needed to be converted into a single row per project. 10 engineered
variables describing timesheets: Percent of hours performed by
`professional' employees as opposed to `technical' employees, Position
of the employee that completed the most hours on each project 3
engineered from invoicing data: 4 from project data: number of projects
completed with each client text analysis of project titles to develop
project classifications

Discretisation was performed by generating a hierarchical dendrogram of
each variable to visualise the clusters.

Limiting the predictive model to a concise set of meaningful variables
reduces noise and improves predictions. Less variables and a simpler
model is easier for stakeholders to understand (Weisberg 2005). followed
by variable importance analysis, variable selection: varialbe importance
aalyses done with linear regression, random forest, cforest to represent
the range of algroithms that were trialled. 11 variables chosen:

trials of selected predictive algorithms. binary classification problem
predicting profit or loss. Complex models should be measured against
simple models that can be built at a fraction of the computational cost.

\begin{itemize}
\tightlist
\item
  Regression - baseline model
\item
  Naive Bayes - baseline model
\item
  Bayesian Network - grabage.
\item
  Random Forest
\item
  Gradient Boosted Trees
\end{itemize}

To compare the models, the area under the receiver operating
characteristic (ROC) curve (AUC) statistic was used. An ROC graph
visualises the curve from which an AUC score is calculated. In this
problem a `positive' is a loss making job. For binary classification,
AUC is a more meaningful statistic than classification accuracy when the
output is a probability that can be applied to the problem. The AUC
indicates model performance across many thresholds while classification
accuracy represents a single threshold.

a model must be first created using a training set. The model can then
make predictions on the test set. this is used to build the ROC A model
that is perfectly classified would have an AUC = 1. A model that
predicts as well as random chance would have an AUC = 0.5. An AUC
between 0.5 and 1 means the model is performing better than random
chance.

In order to compare which models performed significantly better than
others, an adequate sample size of results statistics is required.
Multiple models could be made by using different data in the training
vs.~testing sets, each providing a resulting test statistic (RMSE or
AUC). Initially, 20 models of each method were created in this fashion.
Then a two-sample power calculation was run using the two sets of 20
results to determine the sample size to achieve a statistical power of
0.8 (100 to achieve power of 0.8 between boosted trees and naive bayes)

Missing Data Imputation. All methods except gradient Boosted Trees and
Naive Bayes could not handle missing data. a complete data set allows
for complete sets of predictions from each method, and these predictions
could then be blended to further improve results. The MICE Random Forest
method was chosen for imputation because it has been proven to work well
with complex data sets (Shah et al. 2014). If similar predictive results
were obtained using Boosted Trees imputed data and imputed data, the
imputed data must be reasonable. The imputed data set was then trialed
on the remaining methods and compared to unimputed trials.

Once the best methods were selected, they were blended in numerous ways,
using both simple averaging techniques and sophisticated
machine-learning algorithms. These were compared against individual
models and the best constructs were selected. Six blending methods,
ranging from simple to complex, were tested using predictions from the
top performing individual models. These included simple averaging of the
individual model results, building a Logistic Regression model using the
individual model results only, a Boosted Tree model using individual
model results only, feature weighted linear stacking (FWLS), Random
Forests, and Boosted Trees. A simplified explanation of FWLS is as a
Linear Regression where meta-features as well as model predictions from
individual models are included as explanatory variables. Then, each
meta-feature is interacted with each set of model results (Sill et al.
2009). Feature interaction is performed passively due to the nature of
how trees are built. A split in a node determined by one variable is
conditional upon the preceding split, which was based on another
variable and so on. The simplest method averaged the probability output
(a number between 0 and 1) of all three models. The next two simplest
methods consisted of building a Logistic Regression and Boosted Trees
model from the three prediction model outputs only. Feature Weighted
Linear Stacking (FWLS), Random Forest and Boosted Trees were also tested
as blending methods in a more complex scenario where the predictions
from each model became additional variables to the original explanatory
variables (called meta-features in this context). The model predictions
were interacted with each original variable so that models that
performed more strongly under certain meta-feature states could be
weighted as such.

Again, 100 models were built to achieve a power of 0.8 for the variation
in AUC test statistics across different divisions of training/testing
data. A maximum of five blended models were created from a complete set
of test results (built from the individual models), then a new set of
test results were created for the next five blended models.

Finally the impact of the algorithm on the overall profits of the case
study company was analysed via decision-making scenarios. This analysis
differed from accuracy in predicting profitability, so several blended
models as well as individual models were carried forward for this
analysis. This was done using a profit curve - a chart that plots the
change in profit the company earns on the y-axis vs.~the probability
threshold on the x-axis. A simple approach was taken for this analysis,
where projects with a probability to be a loss making job greater than
the threshold were rejected entirely. Therefore all profits and losses
from jobs above the threshold were discounted. An equation defining the
change in profit as a percentage of the original profit using threshold
rejection is shown below:

\begin{equation}
\Delta\ Overall\ Profit\ \textrm{(\%)} = \frac{\sum_{p = 1}^{N} I(Pr_p < threshold) \cdot Profit_p}{\sum_{p = 1}^{N} Profit_p} \cdot100
\end{equation}

Where

\[
\begin{aligned}
I(\cdot) &= \textrm{the indicator function} \\
N\ &= \textrm{the number of individual projects that are being included in the analysis} \\
Pr_{p} &= \textrm{probability output from the algorithm for project}\ p \textrm{. Values are}\\
& \qquad \textrm{betweeen 0 and 1 where 1 is loss making)} \\
threshold &= \textrm{a chosen value between 0 and 1.}\ \Delta\ Overall\ Profit\ \textrm{is calculated for} \\
& \qquad \textrm{several}\ threshold\ \textrm{values which defines the profit curve} \\
Profit_p &= \textrm{profit for individual project}\ p
\end{aligned}
\]

If the threshold was zero, all jobs were rejected and the profit would
be \$0. If the threshold was 1.0, all jobs were accepted and the profit
would be the same as the profit the company actually experienced since
the data is a sample of historic projects. The aim was to find the
optimal threshold point where saying `no' to a job above that level
would result in higher profits, because jobs that were likely to make a
loss were being rejected. This chart will clarify what percentage of
profit increase the company could expect by integrating the algorithm
into decision-making.

the curve varies with different divisions of the data. Therefore, in
order to understand the uncertainty around the profit curve and to
determine which blended or individual model performed statistically
better than others, a large sample size of curves was required. 100 to
achieve a power of 0.8. which was achieved by repeating 5-fold
training/testing splits. A 95\% confidence interval could also be
determined around the highest point on each curve. The final expected
increase in profit and the percentage of projects to be rejected
presented a clear scenario that the case study business managers could
assess in terms of their business strategy.

\section{4. Prediction Results and
Discussion}\label{prediction-results-and-discussion}

@@

\paragraph{4.1 Individual Models}\label{individual-models}

The predictive formula was re-structured so that the 11 explanatory
variables predicted a new response variable: profitable/unprofitable
projects. The five prediction methods, were initially built without
imputing data, by using the maximum amount of data possible depending on
the method. Boosted Trees and Naive Bayes are able to process data with
missing values, so all data could be input (Ridgeway 2015; Meyer et al.
2014). 100 models of each method were built. The violin plot below
summarises the AUC values produced by each method and the `violins' are
coloured according to whether the distributions significantly vary to
Boosted Trees using a critical value of 0.05.

\begin{figure}[htbp]
\centering
\includegraphics{Consulting_Profitability_Paper_files/figure-latex/reduced_violin-1.pdf}
\caption{Violin plot vertically illustrating the distribution of AUC
values from each of the methods when predicting profit/loss. Subsets of
the data were used for Logistic Regression and Random Forests in order
to provide datasets without missing values.}
\end{figure}

The AUC performance of Logistic Regression and the Random Forest
algorithms cannot be statistically differentiated from Boosted Trees.

Next, data imputation methods were trialed which would make the complete
data set available to Logistic Regression and Random Forests. Again, 100
models were required to achieve a power of 0.8 with respect to Boosted
Trees

\begin{figure}[htbp]
\centering
\includegraphics{Consulting_Profitability_Paper_files/figure-latex/mice_violin-1.pdf}
\caption{Violin plot vertically illustrating the distribution of AUC
values from each of the methods when predicting profit/loss. Each method
was fed the same imputed full dataset.}
\end{figure}

Boosted Trees, Logistic Regression, and Random Forest all performed
significantly better than the baseline algorithm, Naive Bayes, however
none outperformed Boosted Trees.

Testing Boosted Trees, Random Forests, Naive Bayes, Logistic Regression
and Bayesian Networks on the binary classification problem showed that
Boosted Trees, Logistic Regression and Random Forests performed best
(according to AUC). Naive Bayes and Logistic Regression were included as
baseline models and it was expected that the more complex models would
outperform these. Therefore it was surprising that results from 100
Logistic Regression models were not statistically significantly lower
than Boosted Trees.

A possible explanation for this, according to the literature, is that
there was not enough data for the ensemble trees to learn the complex
decision rules at which they excel. Trees tend to overfit the patterns
in a smaller training set. Logistic Regression on the other hand is
capable of only one decision boundary (which does not have to be
parallel to the variable axes) and is not prone to overfitting (Perlich,
Provost, and Simonoff 2003). This may explain Logistic Regression's
comparatively high performance on the case study's small but complex
data set.

Boosted trees, Random Forests and Logistic Regression all had mean AUC
values between 0.755 to 0.764. In summary, the individual binary
classification models performed well above random chance (AUC = 0.5).
Whether the model is worth implementing in the work place is dependent
on the extent to which the algorithm would improve `bottom line' profits
for the business and if the model can affect decisions in practice.

\begin{itemize}
\tightlist
\item
  NB and BN were bad - one sentence
\end{itemize}

\paragraph{4.2 Blended Models}\label{blended-models}

Blended models combine the predictions from each high-performing
individual model to create averaged or `blended' predictions. In this
case, the Random Forest, Boosted Trees, and Logistic Regression models
were chosen. Through variable importance studies, 6 variables were
included as metafeatures. All six methods were compared against the
original Logistic Regression model's AUC distribution (as it was not
statistically different from the ensemble trees)

\begin{figure}[htbp]
\centering
\includegraphics{Consulting_Profitability_Paper_files/figure-latex/blend_violin-1.pdf}
\caption{Violin plot vertically illustrating the distribution of AUC
values from each of the blending methods when predicting profit/loss.
100 models were built for each method.}
\end{figure}

The above plot shows that 4 methods had a higher mean AUC than the
original Logistic Regression model:

\begin{itemize}
\tightlist
\item
  the simple blended Logistic Regression
\item
  the simple average
\item
  and the two Boosted Tree models.
\end{itemize}

The simple averaged method and simple logistic regression could achieve
a statistical power of 0.8 with 150 samples (models), whereas the
boosted trees required thousands. The additional 50 models were run for
the simple logistic regression and simple average to reveal similar
distributions that were indeed statistically significantly higher than
the individual logistic regression AUC distribution. P-values from two
sample t-test were 0.0019 and 0.0021 respectively.

Blended models improved the mean AUC from 0.759 to 0.77, which is an
increase of 1.4\%. It was expected that model blending would improve
predictions because it combined the strengths of individual models with
different theoretical foundations.

The trials of different blending methods demonstrated that again, the
simplest methods worked best with the case study data. In this incident,
averaging the results of the best three single models (Logistic
Regression, Random Forests, and Boosted Trees) or taking a Logistic
Regression of the three models outperformed more complex Boosted Trees,
Random Forests, and Logistic Regression blends that facilitated
interaction of the individual model results and original variables
(meta-features). Simple blended models could perform better than complex
methods if the data is not big enough for the complex models to learn
the more intricate patterns at which they excel. This was observed in
the single model methods. their success in the 2009 Netflix competition
generated some publications. The Netflix data set comprised of almost
3,000,000 observations, so the size of the data could have enabled the
success of complex blending methods such as FWLS (Sill et al. 2009)

\section{5. Business Impact}\label{business-impact}

This chapter first presents the full range of method results in terms of
improvements to the case study's bottom line. A business decision making
scenario was created so that profit curves based on the decision rule
could be built and analysed. From the profit curves, optimal probability
thresholds could be derived for each method (individual and blended
methods). The predictive models output a `probability' between 0 and 1
that each project will be a loss making job (where probability = 1
indicates a loss making job). The question the arises, at what
probability would a decision-maker round the probability up to 1 or down
to 0? And what business decision would then be made? To find the
threshold point for rounding, an experimental business-scenario was
tested. At a given threshold, all projects with probability outputs
above the threshold were considered too risky, and were rejected. All
profits and losses from these projects were forfeited, while the profits
from the remaining jobs (below the threshold) were summed to give a
revised total profit. This profit calculation was made on a range of
thresholds between 0 and 1 at 0.05 increments. The total profits were
plotted for each threshold and joined to make a profit curve

The plot below illustrates the distribution of profit curves for each
method, where the solid lines join the mean values at each threshold
point. The grey ribbon illustrates the 95\% confidence interval for the
profit improvement ratio at each threshold point for the 100 models.

\begin{figure}[htbp]
\centering
\includegraphics{Consulting_Profitability_Paper_files/figure-latex/profit_curve-1.pdf}
\caption{Profit curves summarising results from 100 models of 9 methods:
3 simple blends, 3 complex blends, and the original 3 best methods.}
\end{figure}

Two blended models clearly outperformed the individual models as shown
in the above plot with higher profit ratios as well as tighter
confidence intervals. The simple Logistic Regression blend performed
best with the highest mean profit ratio of 109\% with a standard
deviation of 1.28\%. \%. This means that for the simple blended Logistic
Regression model, if all jobs above the probability threshold 0.6 were
rejected, the profits would increase on average by 9\% in comparison to
historical profits.

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Prediction of Consulting Project Profitability},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  \title{Prediction of Consulting Project Profitability}
    \author[Queensland University of Technology]{Amy Cook, Paul Wu, Kerrie Mengersen\corref{c1}}
   \ead{a21.cook@qut.edu.au} 
   \cortext[c1]{Corresponding Author}
      \address[Queensland University of Technology]{School of Mathematical Sciences, George Street, Brisbane, QLD, 4000}

@@

Clearly state the research question and objectives of the work. Briefly
provide any necessary background to frame the research question.
Concisely summarize the major findings/results. Summary of Key Related
Research This section should include a brief summary of key related
research. Emphasis should be on demonstrating the foundation for the
current investigation. Specifically, the goal is to clearly delineate a
gap or missing link that the current research fills. Authors should
avoid presenting a litany of past research and should focus on prior
work necessary to demonstrate the existence of the research gap
addressed in the manuscript.

@@

Intro - pick out bits from thesis

@@

\begin{itemize}
\item
  amount of losses, sources
\item
  `outside' view, that kind of thing
\end{itemize}

@@

@@

@@

\section{3. Methods}\label{methods}

@@

@@

\begin{itemize}
\tightlist
\item
  description of each of the methods: refer thesis
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Regression, classification, blending, bottom line analysis
\end{itemize}

\section{4. Results and Discussion}\label{results-and-discussion}

@@

\paragraph{4.1 Predictive Analysis}\label{predictive-analysis}

\begin{itemize}
\item
  regression was attempted - failed
\item
  binary classification - results of 5 methods
\item
  blended models - improved results
\end{itemize}

\paragraph{4.2 Bottom line analysis}\label{bottom-line-analysis}

\begin{itemize}
\item
  decision scenario
\item
  profit curves
\end{itemize}

\section{7. Decision Support Tool}\label{decision-support-tool}

\begin{itemize}
\item
  how to generate actions from predictive results
\item
  alternative decision scenarios
\item
  user trust
\end{itemize}

\section{9. Conclusions and Future
Work}\label{conclusions-and-future-work}
